---
title: "Who owns the single family residences in Orange County, NC"
format: 
  html:
    title-block-banner: true
    toc: true
    toc-location: left
    toc-depth: 3
    html-math-method: katex
    code-fold: true
    code-summary: "show the code"
    code-overflow: wrap
    code-tools: true
    number-sections: true
    theme: 
      dark: slate
      light: flatly
    fig-width: 9
    fig-height: 6
editor: visual
---

## Introduction

Let's find out how many residents in Orange County, NC own multiple single-family homes or townhomes!

## Get some data

We need: Owner name and Structure type.

Let's pull in the Orange County parcel data, which includes owner information. It's available [here](ParcelsGDB.zip).

We also need the Orange County Address data, which has the land use code. That's available [here](%22https://web.co.orange.nc.us/gisdownloads/addresses.zip%22).

Finally, we need the municipal boundaries to accurately capture which jurisdictions are relevant. That's available [here](%22https://web.co.orange.nc.us/gisdownloads/city.zip%22).

```{r, output=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(Hmisc)
library(mdbr)
library(knitr)
library(sf)
```

```{r, eval = FALSE}
download.file("https://web.co.orange.nc.us/gisdownloads/ParcelsGDB.zip", destfile = "ParcelsGDB.zip")
unzip("ParcelsGDB.zip")

download.file("https://web.co.orange.nc.us/gisdownloads/addresses.zip", destfile = "addresses.zip")
unzip("addresses.zip")

download.file("https://web.co.orange.nc.us/gisdownloads/city.zip", destfile = "city.zip")
unzip("city.zip")
```

## Read in the data!

```{r}
data <- read_mdb("ParcelsGDB.mdb", table="PARCELVIEW")
addr <- read_sf("addresses.shp")
city <- read_sf("city.shp")
```

## Summarise the data

First, we filter the addresses to only single family-type residences. This includes:

-   Detached single-family

-   Attached single-family

-   Duplex

-   Zero lot line row houses

-   Accessory units

-   Townhouses

-   Manufactured houses

This corresponds to those whose address LBCS Structure codes are in category **1100 other than condominiums (category 1170)** and **single apartments (category 1160)**

```{r, results="asis"}
addr <- addr %>% filter(LBCSSubCod < 1160) %>% filter(LBCSSubCod > 0)
ggplot(data=addr, aes(x=LBCSDesc)) +
  geom_bar()

```

We then join in the actual city boundaries.

```{r}
addr <- st_join(addr,city)
addr$NAME[which(is.na(addr$NAME))] <- "Unincorporated"
```

Now we join the address data to the parcel data, so we can summarise by owner count. We tag owners with "same city" if their mailing address is in same city and/or zip code as the address of the parcel. We tag owners with "in state" if their mailing address is in NC. We tag them as individuals if their name does not include the terms LLC, University, Carolina, INC, Company, Home Trust or Habitat (for Humanity).

```{r}
data$STATE <- gsub(" ", "", data$STATE)
d <- left_join(data,addr,by="PIN") %>% 
  filter(PIN %in% addr$PIN) %>% 
  mutate(owner_name = paste0(OWNER1_FIRST," ",OWNER1_LAST),
         parcel_count = 1,
         samecity_owner = case_when(
           ZIPCODE == ZipCode | CITY == MailingCit ~ 1,
           TRUE ~ 0
         ),
         in_state_owner = case_when(
           STATE == "NC" ~ 1,
           TRUE ~ 0
         ),
         corp_owner = case_when(
           grepl("LLC", owner_name) | grepl("CAROLINA", owner_name) | grepl("HOME TRUST", owner_name) | grepl("UNIVERSITY", owner_name) | grepl("CORP", owner_name) | grepl(" INC ", owner_name) | grepl("INC ", owner_name) | grepl(" INC", owner_name) | grepl("COMPANY", owner_name)| grepl("TOWN", owner_name)| grepl("CITY", owner_name)|  grepl("HABITAT", owner_name) ~ 1,
           TRUE ~ 0
         )
  )

```

### How many people in each city own more than one house there?

OK first, how many **individual/non-corporate owners who live in the same city as their houses** own more than one house in each City in Orange County

```{r, results="asis"}
multi_owners <- d %>%
  group_by(owner_name, corp_owner, samecity_owner, in_state_owner, NAME) %>%
  filter(corp_owner == 0) %>%
  filter(samecity_owner == 1) %>%
  filter(NAME != "DURHAM") %>%
  summarise(owner_count = sum(parcel_count)) %>%
  mutate(owns_multiple = case_when(
    owner_count > 1 ~ "same-city individuals with more than one house",
    TRUE ~ "same-city indivudals with only one house")
  )

kable(table(multi_owners$NAME,multi_owners$owns_multiple))

```

### Aggregate home ownership by owner type

OK Now. Let's see how many **homes** in each city are owned by each of

a\) Same-city individuals with 1 house

b\) Same-city individuals with 2+ houses

c\) Non-local individuals with 1+ houses

d\) Corporations with 1+ houses

e\) government or universities or community home trust with 1+ houses

```{r, results="asis"}
ownership_by_type <- d %>%
  group_by(owner_name, corp_owner, samecity_owner, in_state_owner, NAME) %>%
  filter(NAME != "DURHAM") %>%
  summarise(owner_count = sum(parcel_count)) %>%
  mutate(owns_multiple = case_when(
    owner_count > 1 ~ "same-city individuals with more than one house",
    TRUE ~ "same-city indivudals with only one house")
  ) %>%
  mutate(
    owner_type = case_when(
      corp_owner == 0 & samecity_owner == 1 & owner_count ==1 ~ "a. same city individual with 1 house",
      corp_owner == 0 & samecity_owner == 1 & owner_count >1 ~  "b. same city individual with 2+ house",
      corp_owner == 0 & samecity_owner == 0 ~ "c. nonresident individual owner",
      corp_owner == 1 & !(grepl("UNIVERSITY",owner_name) | grepl("TOWN", owner_name) | grepl("CAROLINA", owner_name)| grepl("HOME TRUST", owner_name)| grepl("HABITAT", owner_name)) ~ "d. corp",
      corp_owner == 1 & (grepl("UNIVERSITY",owner_name)| grepl("TOWN", owner_name) | grepl("CAROLINA", owner_name)| grepl("HOME TRUST", owner_name) | grepl("HABITAT", owner_name)) ~ "e. university/govt/home trust"                    )
  ) %>%
  group_by(owner_type,NAME) %>%
  summarise(houses_owned = sum(owner_count)) %>% filter(!(NAME %in% c("Unincorporated")))

ggplot(ownership_by_type, aes(fill=owner_type, y= houses_owned, x=NAME)) +
  geom_bar(position="stack", stat="identity") + geom_text(aes(label=houses_owned), position = "stack")
```
